\Large {\bf rt1d: User's Manual}

### 2012-04-15 doc/manual/rt1d_manual.tex
General cleanup + test problem 0 + recombination cooling bug fix + other stuff.

\usepackage{amsmath, amsthm, amssymb}
\numberwithin{equation}{section}
\usepackage{graphicx}
\usepackage{pslatex}
\usepackage{natbib}
\usepackage[pdfborder={0 0 0}]{hyperref}
\usepackage[all]{hypcap}

%%% Beginning of Document

\begin{document}

@@

\author{Jordan Mirocha}

\title{\Large {\bf rt1d: User's Manual}}
\date{Last Updated: \today}
\maketitle

@@

%%%
%% Introduction
%%%
\section{Introduction}
Let's see what happens to the gas around stars and black holes, shall we?

%%% 
%% Methods
%%%
\section{Methods}
\subsection{The Rate Equations}
In general, the chemical and thermal evolution of gas surrounding a radiation
source is governed by a set of differential equations describing the number
densities of all ions and the temperature of the gas. Assuming a medium
consisting of hydrogen and helium only, we first solve for the abundances of
each ion via
\begin{align}
    \frac{d \nHII}{dt} & = (\gHI + \gamma_{\HI} + \bHI \nel) \nHI - \aHII \nel \nHII   \label{eq:HIIRateEquation} \\ 
    \frac{d \nHeII}{dt} & = (\gHeI + \gamma_{\HeI} + \bHeI \nel) \nHeI \nonumber + \aHeIII \nel \nHeIII \\  & - (\bHeII + \aHeII + \xiHeII) \nel \nHeII \label{eq:HeIIRateEquation} \\ 
    \frac{d \nHeIII}{dt} & = (\gHeII + \gamma_{\HeII} + \bHeII \nel) \nHeII  - \aHeIII \nel \nHeIII . \label{eq:HeIIIRateEquation}
\end{align}
Each of these equations represents the balance between ionizations of species
$\HI$, $\HeI$, and $\HeII$, and recombinations of $\HII$, $\HeII$, and
$\HeIII$. Associating the index $i$ with absorbing species, $i = \HI, \HeI,
\HeII$, we define $\Gamma_i$ as the photo-ionization rate coefficient,
$\gamma_i$ as the secondary ionization rate coefficient, $\alpha_i$ ($\xi_i$)
as the case-B (dielectric) recombination rate coefficients, $\beta_i$ as the
collisional ionization rate coefficients, and $\nel = \nHII + \nHeII +
2\nHeIII$ as the number density of electrons.

@@

% Photon-conservation
\subsection{Computing the Ionization and Heating Rates}
The most critical aspect of propagating the radiation field in our 1D
simulations is computing the ionization ($\Gamma_i$, $\gamma_{i}$) and heating
($\Heat_i$) rate coefficients accurately. In order to directly relate our
results to fully 3-dimensional radiative transfer calculations we have chosen
to adopt a photon-conserving (PC) algorithm nearly identical to those employed
by several widely used codes, like \textit{C$^2$Ray} \citep{Mellema2006}, and
\textit{Enzo} \citep{Wise2011}. Our code is able to compute $\Gamma_i$,
$\gamma_{i}$, and $\Heat_i$ in a non-photon-conserving (NPC) fashion as well,
to enable comparison with previous 1D work such as \citet{Thomas2008}. The two
formalisms are equivalent in the limit of very optically thin cells, a
condition that can be met easily in 1D calculations but is rarely
computationally feasible in 3D. For NPC methods, if the optical depth of an
individual cell is substantial, the number of ionizations in that cell will
\textit{not} equal the number of photons absorbed for that cell, i.e. photon
number will not be conserved. This problem was remedied by \citet{Abel1999},
who inferred the number of photo-ionizations of species $i$ in a cell from the
radiation incident upon it and its optical depth,
\begin{equation}
    \Delta \tau_{i,\nu} = n_i \sigma_{i,\nu} \Delta r .
\end{equation}    
It is most straightforward to imagine our 1D grid as a collection of
concentric spherical shells, each having thickness $\Delta r$ and volume
$V_{\sh}(r) = 4 \pi [(r + \Delta r)^3 - r^3] / 3$. The ionization and heating
rates can then be related to the number of absorptions in any given shell
(thus preserving photon number), as
\begin{align}
    \Gamma_i & = A_i \int_{\nu_i}^{\infty} I_{\nu} e^{-\tau_{\nu}} \left(1 - e^{-\Delta \tau_{i,\nu}}\right) \frac{d\nu}{h\nu} \label{eq:PhotoIonizationRate} \\
    \gamma_{ij} & = A_j \int_{\nu_j}^{\infty} \left(\frac{\nu - \nu_j}{\nu_i}\right) I_{\nu} e^{-\tau_{\nu}} \left(1 - e^{-\Delta \tau_{j,\nu}}\right) \frac{d\nu}{h\nu} \label{eq:SecondaryIonizationRate} \\
    \Heat_i & = A_i \int_{\nu_i}^{\infty} (\nu - \nu_i) I_{\nu} e^{-\tau_{\nu}} \left(1 - e^{-\Delta \tau_{i,\nu}}\right) \frac{d\nu}{\nu} , \label{eq:HeatingRate}
\end{align}    
where we've defined the normalization constant, $A_i \equiv
L_{\mathrm{bol}}/n_i V_{\sh}(r)$, and denote the ionization threshold energy
for species $i$ as $h\nu_i$.

@@

For simplicity, our current treatment neglects a few physical processes that
are cosmological in origin, or simply do not rely on the radiation field
directly. These include cooling via free-free emission and hydrogen ionization
due to helium recombination photons (which depend on the gas kinetic
temperature and electron density), and cosmological effects such as Hubble
cooling, Compton cooling off cosmic microwave background (CMB) photons, and
photo-ionization by Wien-tail CMB photons (which depend on kinetic
temperature, redshift, and the Hubble parameter). Our optimized SEDs
(\S\ref{sec:Results}) will be unchanged upon including such physics in later
work.

\subsection{Additional Complexity: Energy Dependent Secondary Electron Treatment}
In the previous section (and in all of Mirocha et al. 2012), we considered the asymptotic limit of \citet{Shull1985}, in which the fractional energy deposition of secondary electrons as heat, ionization, and excitation depends only on the hydrogen ionized fraction.  \citet{Furlanetto2010} updated this work, providing lookup tables of deposition fraction as a function of both ionized fraction and electron energy.  These effects are most important at $E \lesssim 10^2$ eV.

To accommodate this model, we must redefine the quantities $\Phi_i$ and $\Psi_i$.  We will use a tilde to denote the new quantities related to secondary ionization, 
\begin{align}
    \widetilde{\Phi}_{ij} & \equiv \int_{\nu_j}^{\infty} \fionE I_{\nu} e^{-\tau_{\nu}} \frac{d\nu}{h\nu} \\
    \widetilde{\Psi}_{ij} & \equiv \int_{\nu_j}^{\infty} \fionE I_{\nu} e^{-\tau_{\nu}} d\nu ,
\end{align}
and a hat to denote the new quantities related to photo-electric heating,
\begin{align}
    \widehat{\Phi}_i & \equiv \int_{\nu_i}^{\infty} \fheatE I_{\nu} e^{-\tau_{\nu}} \frac{d\nu}{h\nu} \\
    \widehat{\Psi}_i & \equiv \int_{\nu_i}^{\infty} \fheatE I_{\nu} e^{-\tau_{\nu}} d\nu .
\end{align}
Now, our ionization and heating coefficients read
\begin{align}
    \Gamma_i & = A_i \left[\Phi_i(\tau_{\nu}) - \Phi_i(\tau_{i,\nu}^{\prime}) \right] \label{eq:Gamma_PhiPsi_complex} \\
    \gamma_{ij} & = \frac{A_j}{h\nu_i} \left\{\widetilde{\Psi}_{ij}(\tau_{\nu}) - \widetilde{\Psi}_{ij}(\tau_{j,\nu}^{\prime}) - h \nu_j \left[\widetilde{\Phi}_{ij}(\tau_{\nu}) - \widetilde{\Phi}_{ij}(\tau_{j,\nu}^{\prime}) \right] \right\}  \label{eq:gamma_PhiPsi_complex} \\
    \Heat_i & = A_i \left\{\widehat{\Psi}_i(\tau_{\nu}) - \widehat{\Psi}_i(\tau_{i,\nu}^{\prime}) - h\nu_i \left[\widehat{\Phi}_i(\tau_{\nu}) - \widehat{\Phi}_i(\tau_{i,\nu}^{\prime}) \right] \right\} \label{eq:Heat_PhiPsi_complex} .  
\end{align}

@@

\subsection{\texttt{InitializeGrid}}
\subsection{\texttt{InitializeIntegralTables}}
\subsection{\texttt{RadiationSource}}
\subsection{\texttt{Radiate}}
\subsection{\texttt{SolveRateEquations}}
\subsection{\texttt{Interpolate}}
\subsection{\texttt{WriteData}}

%%% 
%% Parameters
%%%
\section{Parameters}

@@

\item [\texttt{ProblemType = 2.1}] Same as \texttt{ProblemType = 2} for everything, except we sample the continuous blackbody spectrum with the four energy groups used by \citet{Wise2011}

\item [\texttt{ProblemType = 2.2}] Same as \texttt{ProblemType = 2} for everything, except we sample the continuous blackbody spectrum with the four energy groups used by Mirocha et al. 2012.

\item [\texttt{ProblemType = 3.1}] Test problem \# 3 from \cite{Iliev2006} -- I-front trapping in a dense clump and the formation of a shadow.  The setup is the same as in \texttt{ProblemType = 2.1}, except a 1D clump is initialized at $x = 0.76 L_{\mathrm{box}}$, overdensity $\delta = 200$, radius $0.12$kpc, and temperature $T = 40$K, and the radiation field is plane-parallel.
\end{description}

@@

\subsection{Physics Parameters}

@@

@@

%How to change the code to allow for this?
%Define main loop in RT1D.py as Shine.  Insert if __name__ == `main' stuff?
%Would be nice to have an executable...
%Move RT1D.py to \$RT1D/bin, import it from \$RT1D/rt1d/__init__.py

@@

%In this work, we use the formulae in Appendix B of \cite{Fukugita1994} to
%compute the values of $\alpha_i$, $\beta_i$, and $\xi_i$. As you might expect, collisional ionization and recombination depend solely on the temperature of the gas.  For example, collisional ionization of hydrogen (from $n = 1$) is governed by
%\begin{equation}
%    \beta_{1,\HI} = 5.85 \times 10^{-11} T^{1/2} \left[1 + \left(\frac{T}{10^5 \mathrm{K}}\right)^{1/2}\right]^{-1} \mathrm{exp}\left[-1.578 \times \left(\frac{T}{10^5 \mathrm{K}}\right)^{-1}\right] \mathrm{cm^3 s^{-1}}
%\end{equation}
%and recombinations (to $n \ge 1$) by
%\begin{equation}
%    \alpha_{\HII} = 6.28 \times 10^{-11} T^{-1/2} \left(\frac{T}{10^3\mathrm{K}}\right)^{-1/5} \left[1 + \left(\frac{T}{10^6\mathrm{K}}\right)^{0.7}\right]^{-1} \mathrm{cm^3s^{-1}}
%\end{equation}

>>>>

\usepackage{amsmath, amsthm, amssymb}
\numberwithin{equation}{section}
\usepackage{graphicx}
\usepackage{pslatex}
\usepackage{epsfig}
\usepackage{natbib}
\usepackage[pdfborder={0 0 0}]{hyperref}
\usepackage[all]{hypcap}

@@

%%% Beginning of Document
\begin{document}
\author{Jordan Mirocha}	
\title{\Large {\bf rt1d: User's Manual}}
\date{Last Updated: \today}
\maketitle

@@

%%%
%% Introduction
%%%
\section{Introduction}
Let's see what happens to the gas around stars and black holes, shall we? The
primary reference for this code is
\href{http://arxiv.org/abs/1204.1944}{arxiv:1204.1944} -- much of the text has
been extracted from there.

%%% 
%% Methods
%%%
\section{Methods}

% Rate Equations
\subsection{The Rate Equations}
In general, the chemical and thermal evolution of gas surrounding a radiation
source is governed by a set of differential equations describing the number
densities of all ions and the temperature of the gas. Assuming a medium
consisting of hydrogen and helium only, we first solve for the abundances of
each ion via
\begin{align}
    \frac{d \nHII}{dt} & = (\gHI + \gamma_{\HI} + \bHI \nel) \nHI - \aHII \nel \nHII   \label{eq:HIIRateEquation} \\ 
    \frac{d \nHeII}{dt} & = (\gHeI + \gamma_{\HeI} + \bHeI \nel) \nHeI \nonumber + \aHeIII \nel \nHeIII \\  & - (\bHeII + \aHeII + \xiHeII) \nel \nHeII \label{eq:HeIIRateEquation} \\ 
    \frac{d \nHeIII}{dt} & = (\gHeII + \gamma_{\HeII} + \bHeII \nel) \nHeII  - \aHeIII \nel \nHeIII . \label{eq:HeIIIRateEquation}
\end{align}
Each of these equations represents the balance between ionizations of species
$\HI$, $\HeI$, and $\HeII$, and recombinations of $\HII$, $\HeII$, and
$\HeIII$. Associating the index $i$ with absorbing species, $i = \HI, \HeI,
\HeII$, we define $\Gamma_i$ as the photo-ionization rate coefficient,
$\gamma_i$ as the secondary ionization rate coefficient, $\alpha_i$ ($\xi_i$)
as the case-B (dielectric) recombination rate coefficients, $\beta_i$ as the
collisional ionization rate coefficients, and $\nel = \nHII + \nHeII +
2\nHeIII$ as the number density of electrons.

@@

% Ionization and Heating Rates
\subsection{Computing the Ionization and Heating Rates}
The most critical aspect of propagating the radiation field in our 1D
simulations is computing the ionization ($\Gamma_i$, $\gamma_{i}$) and heating
($\Heat_i$) rate coefficients accurately. In order to directly relate our
results to fully 3-dimensional radiative transfer calculations we have chosen
to adopt a photon-conserving (PC) algorithm nearly identical to those employed
by several widely used codes, like \textit{C$^2$Ray} \citep{Mellema2006}, and
\textit{Enzo} \citep{Wise2011}. Our code is able to compute $\Gamma_i$,
$\gamma_{i}$, and $\Heat_i$ in a non-photon-conserving (NPC) fashion as well,
to enable comparison with previous 1D work such as \citet{Thomas2008}. The two
formalisms are equivalent in the limit of very optically thin cells, a
condition that can be met easily in 1D calculations but is rarely
computationally feasible in 3D. For NPC methods, if the optical depth of an
individual cell is substantial, the number of ionizations in that cell will
\textit{not} equal the number of photons absorbed for that cell, i.e. photon
number will not be conserved. This problem was remedied by \citet{Abel1999},
who inferred the number of photo-ionizations of species $i$ in a cell from the
radiation incident upon it and its optical depth,
\begin{equation}
    \Delta \tau_{i,\nu} = n_i \sigma_{i,\nu} \Delta r .
\end{equation}    
It is most straightforward to imagine our 1D grid as a collection of
concentric spherical shells, each having thickness $\Delta r$ and volume
$V_{\sh}(r) = 4 \pi [(r + \Delta r)^3 - r^3] / 3$ (i.e. radii correspond to
cell interfaces). The ionization and heating rates can then be related to the
number of absorptions in any given shell (thus preserving photon number), as
\begin{align}
    \Gamma_i & = A_i \int_{\nu_i}^{\infty} I_{\nu} e^{-\tau_{\nu}} \left(1 - e^{-\Delta \tau_{i,\nu}}\right) \frac{d\nu}{h\nu} \label{eq:PhotoIonizationRate} \\
    \gamma_{ij} & = A_j \int_{\nu_j}^{\infty} \left(\frac{\nu - \nu_j}{\nu_i}\right) I_{\nu} e^{-\tau_{\nu}} \left(1 - e^{-\Delta \tau_{j,\nu}}\right) \frac{d\nu}{h\nu} \label{eq:SecondaryIonizationRate} \\
    \Heat_i & = A_i \int_{\nu_i}^{\infty} (\nu - \nu_i) I_{\nu} e^{-\tau_{\nu}} \left(1 - e^{-\Delta \tau_{i,\nu}}\right) \frac{d\nu}{\nu} , \label{eq:HeatingRate}
\end{align}    
where we've defined the normalization constant, $A_i \equiv
L_{\mathrm{bol}}/n_i V_{\sh}(r)$, and denote the ionization threshold energy
for species $i$ as $h\nu_i$.

@@

% Secondary Electrons
\subsection{Additional Complexity: Energy Dependent Secondary Electron Treatment}
In the previous section (and in all of Mirocha et al. 2012), we considered the asymptotic limit of \citet{Shull1985}, in which the fractional energy deposition of secondary electrons as heat, ionization, and excitation depends only on the hydrogen ionized fraction.  \citet{Furlanetto2010} updated this work, providing lookup tables of deposition fraction as a function of both ionized fraction and electron energy.  These effects are most important at $E \lesssim 10^2$ eV.

To accommodate this model, we must redefine the quantities $\Phi_i$ and $\Psi_i$.  We will use a tilde to denote the new quantities related to secondary ionization, 
\begin{align}
    \widetilde{\Phi}_{ij} & \equiv \int_{\nu_j}^{\infty} \fionE I_{\nu} e^{-\tau_{\nu}} \frac{d\nu}{h\nu} \\
    \widetilde{\Psi}_{ij} & \equiv \int_{\nu_j}^{\infty} \fionE I_{\nu} e^{-\tau_{\nu}} d\nu ,
\end{align}
and a hat to denote the new quantities related to photo-electric heating,
\begin{align}
    \widehat{\Phi}_i & \equiv \int_{\nu_i}^{\infty} \fheatE I_{\nu} e^{-\tau_{\nu}} \frac{d\nu}{h\nu} \\
    \widehat{\Psi}_i & \equiv \int_{\nu_i}^{\infty} \fheatE I_{\nu} e^{-\tau_{\nu}} d\nu .
\end{align}
Now, our ionization and heating coefficients read
\begin{align}
    \Gamma_i & = A_i \left[\Phi_i(\tau_{\nu}) - \Phi_i(\tau_{i,\nu}^{\prime}) \right] \label{eq:Gamma_PhiPsi_complex} \\
    \gamma_{ij} & = \frac{A_j}{h\nu_i} \left\{\widetilde{\Psi}_{ij}(\tau_{\nu}) - \widetilde{\Psi}_{ij}(\tau_{j,\nu}^{\prime}) - h \nu_j \left[\widetilde{\Phi}_{ij}(\tau_{\nu}) - \widetilde{\Phi}_{ij}(\tau_{j,\nu}^{\prime}) \right] \right\}  \label{eq:gamma_PhiPsi_complex} \\
    \Heat_i & = A_i \left\{\widehat{\Psi}_i(\tau_{\nu}) - \widehat{\Psi}_i(\tau_{i,\nu}^{\prime}) - h\nu_i \left[\widehat{\Phi}_i(\tau_{\nu}) - \widehat{\Phi}_i(\tau_{i,\nu}^{\prime}) \right] \right\} \label{eq:Heat_PhiPsi_complex} .  
\end{align}    
Various methods for handling secondary electrons are compared for a $10^5$ K blackbody source of $\dot{Q} = 5 \times 10^{48}$ ionizing photons $s^{-1}$ in Figure \ref{fig:SecondaryIonizationMethods}.

% Secondary electron methods
\begin{figure*}[htbp!]
\begin{center}
\includegraphics[width=0.98\textwidth]{figures/secondary_ionization_methods.eps}
\caption{Comparison of different methods for secondary ionization.  See \S\ref{sec:PhysicsParameters} for details on setting the secondary ionization method.}
\label{fig:SecondaryIonizationMethods}
\end{center}
\end{figure*}

\subsection{Time-stepping Techniques}

\subsection{Parallelism}

@@

\subsection{\texttt{InitializeGrid}}
Initialize the temperature and ion densities for each cell in the grid.
\subsection{\texttt{InitializeIntegralTables}}
If \texttt{IntegralTabulation} = 1 (necessary when \texttt{DiscreteSpectrum} = 0), tabulates the quantities $\Phi_i$ and $\Psi_i$.  If \texttt{MultiSpecies} = 1, tabulation occurs for hydrogen and helium ions, and if \texttt{SecondaryIonization} $\ge 2$, also tabulate $\widetilde{\Phi}_i$, $\widetilde{\Psi}_i$, $\widehat{\Phi}_i$, and $\widehat{\Psi}_i$, adding an additional dimension for $x_{\mathrm{HII}}$.
\subsection{\texttt{RadiationSource}}
Initialize the radiation source.  Most importantly the \texttt{Spectrum} function, which returns the fraction of the bolometric luminosity emitted at energy $E$ (only argument).
\subsection{\texttt{Radiate}}
Calls \texttt{EvolvePhotons} routines, which loop over grid, call solver, and update ion densities and the temperature in each cell.
\subsection{\texttt{SolveRateEquations}}
ODE solver.
\subsection{\texttt{Interpolate}}
Performs linear, bi-linear, tri-linear, and eventually quad-linear interpolation on integral tables to extract ionization and heating coefficients as a function of column density.
\subsection{\texttt{WriteData}}
Write out data to HDF5.  Would be easy to add support for ASCII write-out.

%%% 
%% Parameters
%%%
\section{Parameters} \label{sec:Parameters}

@@

\item [\texttt{ProblemType = 2.1}] Same as \texttt{ProblemType = 2}, except we sample the continuous blackbody spectrum with the four energy groups used by \citet{Wise2011}.

\item [\texttt{ProblemType = 2.2}] Same as \texttt{ProblemType = 2}, except we sample the continuous blackbody spectrum with the four energy groups used by Mirocha et al. (2012).

\item [\texttt{ProblemType = 3}] Test problem \# 3 from \cite{Iliev2006} -- I-front trapping in a dense clump and the formation of a shadow.  The setup is the same as in \texttt{ProblemType = 2}, except a 1D clump is initialized at $x = 0.76 L_{\mathrm{box}}$, overdensity $\delta = 200$, radius $0.12$kpc, and temperature $T = 40$K, and the radiation field is plane-parallel.

\item [\texttt{ProblemType = 3.1}] Same as \texttt{ProblemType = 3}, except we sample the continuous blackbody spectrum with the four energy groups used by \citet{Wise2011}.

@@

\subsection{Physics Parameters} \label{sec:PhysicsParameters}

@@

%%%
%% Test Problems
%%%
\section{Test Problems}
In this section we focus on test problems 1-3 of the Radiative Transfer Comparison Project \citep[hereafter RT06;][]{Iliev2006}.

\subsection{Test \#0}
Single-zone ionization, heating, and cooling.

\begin{figure*}[htbp!]
\begin{center}
\includegraphics[width=0.98\textwidth]{figures/rt06_0.eps}
\caption{Convergence test for RT06 \#1.}
\label{fig:RT06_1}
\end{center}
\end{figure*}

\subsection{Test \#1}

\begin{figure*}[htbp!]
\begin{center}
\includegraphics[width=0.98\textwidth]{figures/rt06_1_convergence.eps}
\caption{Convergence test for RT06 \#1.}
\label{fig:RT06_1}
\end{center}
\end{figure*}

\subsection{Test \#2}

\begin{figure*}[htbp!]
\begin{center}
\includegraphics[width=0.98\textwidth]{figures/rt06_2.eps}
\caption{RT06 \#3.}
\label{fig:RT06_2}
\end{center}
\end{figure*}

\subsection{Test \#3}

\begin{figure*}[htbp!]
\begin{center}
\includegraphics[width=0.98\textwidth]{figures/rt06_3.eps}
\caption{RT06 \#3.}
\label{fig:RT06_3}
\end{center}
\end{figure*}

@@

@@
